Modelling the Lymphatic Metastatic Progression Pathways of OPSCC \\ from Multi-Institutional Datasets

### 2024-02-05 src/tex/ms.tex
[showyourwork] overleaf sync

% Define document class
\documentclass[twocolumn]{article}
\usepackage[a4paper, total={190mm, 270mm}]{geometry}
\usepackage{showyourwork}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{wasysym}
\usepackage{xcolor}
\usepackage{nicefrac}
\usepackage{multirow}
\usepackage[affil-it]{authblk}
\usepackage{graphicx}
\usepackage{biblatex}
\addbibresource{bib.bib}

@@

            We extend a previously published \gls{hmm}, which models the \glspl{lnl} (I, II, III, IV, V, and VII) as hidden binary \glspl{rv}. Each represents a patient's true state of lymphatic involvement. Clinical involvement at diagnosis represents the observed binary \glspl{rv} linked to the true state via sensitivity and specificity. The primary tumor and the hidden \glspl{rv} are connected in a graph. Each edge represents the conditional probability of metastatic spread per abstract time-step, given disease at the edge's starting node. To learn these probabilities, we draw \acrlong{mcmc} samples from the likelihood of a dataset (\variable{output/num_patients.txt}\gls{opscc} patients) from three institutions. We compute the model evidence using \acrlong{ti} for different graphs to determine which describes the data best.

@@

\begin{table}
\centering
\begin{tabular}{|cc|cc|}
    \hline
    & & $X_2=0$ & $X_2=1$ \\
    \hline
    \multirow{2}{*}{$X_3$} & $=0$ & $1-b_3$ & $(1-b_3)(1-t_{23})$ \\
    & $=1$ & $b_3$ & $1-b_3-t_{23}+b_3 t_{23}$ \\
    \hline
\end{tabular}
\caption{$P\left( X_3 \mid X_2 \right)$ for all four possible combinations of the states of $X_3$ and $X_2$.}
\end{table}

@@

\begin{table}
\centering
\begin{tabular}{|cc|cc|}
    \hline
    & & $X=0$ & $X=1$ \\
    \hline
    \multirow{2}{*}{$Z$} & $=0$ & $s_P$ & $1-s_N$ \\
    & $=1$ & $1-s_P$ & $s_N$ \\
    \hline
\end{tabular}
\caption{$P\left( Z \mid X \right)$ for all possible values of $X$ and $Z$.}
\end{table}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.6\linewidth]{figures/extended_graph.png}
    \caption{Extended \gls{dag} representing different possible spread graphs underlying the \gls{hmm}. As in \cref{fig:graph_with_obs}, red arcs are parametrized with probabilities of spread from the tumor (red circle) to the \glspl{lnl} (blue circles). These red arcs, together with the blue arcs fron \gls{lnl} to \gls{lnl}, make up the \emph{base graph}. One after the other, each of the green arcs was added to the base graph. Subsequently, the performance of the resulting models in terms of its \gls{bic} was compared to the base graph to assess whether the additional edge should be kept in the \emph{winning graph} or not. \label{fig:extended_graph}}
\end{figure}

@@

\begin{figure}[ht]
    \script{thermo_int.py}
    \centering
    \includegraphics[width=\linewidth]{figures/thermo_int.png}
    \caption{The expected log-accuracy under the power posterior plotted against the value of the inverse temperature $\beta$. These values were computed during a TI run for the loosing graph model (red line), as well as the winning graph model (green line). Note that the scaling of the x-axis is chosen such that the 64 points on the temperature ladder appear evenly spaced. This is to stress how the log-accuracy develops in the range from $\beta=0$ to around $\beta=0.1$. The area under these two curves yields an approximation of the log-evidence of the respective model. It is barely visible that due to the simplicity of the loosing graph model its accuracy increases earlier and faster that the winning graph model's. However, already before $\beta=0.031$ the model based on the winning graph outperforms the loosing graph model. Ultimately, the winning graph's log-evidence is better than the loosing graph model's.}
    \label{fig:thermo_int}
\end{figure}

@@

\begin{table}[h]
\centering
\begin{tabular}{|c|c|c|}
    \hline
    $K_\text{1v2}$ & $\ln{K_\text{1v2}}$ & support for $\mathcal{M}_1$ \\
    \hline
    $< 10^0$ & $< 0$ & negative (supports $\mathcal{M}_2$) \\
    $10^0$ to $10^{\nicefrac{1}{2}}$ & 0 to 1.15 & barely worth a mention \\
    $10^{\nicefrac{1}{2}}$ to $10^1$ & 1.15 to 2.3 & substantial \\
    $10^1$ to $10^{\nicefrac{3}{2}}$ & 2.3 to 3.45 & strong \\
    $10^{\nicefrac{3}{2}}$ to $10^2$ & 3.45 to 4.6 & very strong \\
    $> 10^2$ & $> 4.6$ & decisive \\
    \hline
\end{tabular}
\caption{Interpretation of Bayes factors and their natural logarithms in terms of their support for or against one of the two compared models as introduced by \cite{jeffreys_theory_1998}.}
\label{table:bayes_factor}
\end{table}

@@

\begin{table}[h]
\centering
\begin{tabular}{|l|rr|}
    \hline
    \textbf{Modality} & \textbf{Specificity} & \textbf{Sensitivity} \\
    \hline
    \acrshort{ct} & 76\% & 81\% \\
    \acrshort{pet} & 86\% & 79\% \\
    \gls{mri} & 63\% & 81\% \\
    \acrshort{fna} & 98\% & 80\% \\
    Pathology & $\approx$ 100\% & $\approx$ 100\% \\
    \hline
\end{tabular}
\caption{Literature sensitivity and specificity values that we used to infer the most likely involvement for a patient when multiple diagnostic modalities reported conflicting nodal involvement \cite{de_bondt_detection_2007,kyzas_18f-fluorodeoxyglucose_2008}.}
\label{table:sens_spec}
\end{table}

@@

\begin{table}[h]
\centering
\begin{tabular}{|ccccccrrrr|}
    \hline
    \multicolumn{6}{|c}{\textbf{LNL involvement} } & \multicolumn{4}{c|}{\textbf{T-category} } \\
    I & II & III & IV & V & VII & \multicolumn{2}{c}{\textbf{early} } & \multicolumn{2}{c|}{\textbf{advanced} } \\
    \hline
    \variable{output/data_table.tex} \\
    \hline
\end{tabular}
\caption{Prevalence of involvement patterns in the multi-centric dataset. An involvement pattern is characterized by the state of the six \glspl{lnl}: A red dot means the \gls{lnl} was reported to be metastatic, a green dot means it was determined to be healthy and a question mark means that the prevalence was marginalized over the state of this \gls{lnl}.}
\label{table:data_prevalence}
\end{table}

@@

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\linewidth]{figures/ranking.pdf}
    \caption{Visual ranking of the investigated graphs w.r.t. their model evidence, computed via thermodynamic integration. Not shown are the two graphs where the arc from \gls{lnl} III to IV was removed. Their respective model evidence fell below $-1528$ and the two graphs would appear far left in the figure. In the bottom left corner, we provide a visual reference in analogy to \cref{table:bayes_factor}: E.g., any difference in the model evidence shorter than the first of the three rulers indicates that the improvement is ``barely worth a mention''. \label{fig:ranking}}
\end{figure*}

@@

\begin{figure}
    \script{wg_III_risks.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/wg_III_risks.png}
        \caption{Histograms over the risk for microscopic involvement in \gls{lnl} III, given that a patient presents as clinically N0 (green), or given that \gls{lnl} II shows clinical involvement (blue). The top panel shows these risks for early T-category, the bottom panel for late. \label{fig:wg_III_risks}}
    \end{centering}
\end{figure}

@@

@@

\begin{figure}
    \script{wg_I_risks.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/wg_I_risks.png}
        \caption{Distributions over predicted risk for involvement in \gls{lnl} I, given different clinical diagnosis scenarios: For N0 patients (green), patients with macroscopic involvement in \gls{lnl} II (blue), and for the case where the \glspl{lnl} II, III, and IV show involvement. The top panel shows these risks for early T-category and the bottom row for advanced T-category. \label{fig:wg_I_risks}}
    \end{centering}
\end{figure}

@@

\subsection{Implications for Elective\\Nodal Treatment} 
Risk predictions obtained by the model may be used to design clinical trials on volume-deescalated treatment of \gls{opscc}. In the context of radiotherapy, this corresponds to excluding \glspl{lnl} from the \gls{ctv-n}, which are irradiated according to the current guidelines. The list below should be seen as a summary of \cref{subsec:results:risk_prediction} and the limitations discussed in \cref{subsec:disc:limitation} should be taken into account in its interpretation. Assuming that one accepts approximately a 5\% risk of occult metastases per \gls{lnl}, the statistical model presented in this paper would suggest to:
\begin{itemize}
    \item[$\bullet$] Irradiate level II for all patients.
    \item[$\bullet$] Irradiate level III for most patients. Only for clinically N0, early T-category patients, not irradiating level III could be considered. 
    \item[$\bullet$] Exclude level IV from the \gls{ctv-n} for patients with clinically negative level III. For advanced T-category patients with involvement of level III, level IV should be irradiated.
    \item[$\bullet$] Exclude level V from the \gls{ctv-n} for most patients. Only for patients with extensive involvement of levels II, III, and IV, irradiation of level V can be considered.
    \item[$\bullet$] Exclude level I from the \gls{ctv-n} for early T-category patients with limited metastatic disease. For advanced T-category patients with extensive nodal involvement of levels II, III, and IV, level I should be irradiated (see also the limitations discussed in \cref{subsec:disc:limitation}).
    \item[$\bullet$] Exclude level VII in all patients, unless it is clinically involved.
\end{itemize}

@@

@@

The patient data detailing lymphatic involvement is publicly available in the form of CSV tables in the GitHub repository \href{https://github.com/rmnldwg/lydata}{\texttt{rmnldwg/lydata}}.

In another GitHub repository, \href{https://github.com/rmnldwg/lynference}{\texttt{rmnldwg/lynference}}, we define the experiments based on different graph structures. These experiments are reproducible via the tool \href{https://dvc.org}{DVC} and their locally computed results are uploaded to an Azure blob storage container.

Lastly, the \LaTeX\ source code, all Python scripts to generate the figures and tables in this work, and the pipeline definition to build the document (using \showyourwork \cite{luger_mapping_2021}) are made public in the GitHub repository \href{https://github.com/rmnldwg/graph-extension-paper}{\texttt{rmnldwg/graph-extension-paper}}. To rebuild it, the \showyourwork tool pulls the patient data from the \href{https://github.com/rmnldwg/lynference}{\texttt{rmnldwg/lynference}} repository and the trained models in the form of \gls{mcmc} samples from the Azure blob storage container, then runs the Python scripts and compiles the \LaTeX\ article.

Detailed instructions on how to reproduce the entire pipeline, including the individual experiments, are gladly given upon request. They may also be found in the respective repository's \texttt{README.md} file.

>>>>

% Define document class
\documentclass[twocolumn]{article}
\usepackage[a4paper, total={190mm, 270mm}]{geometry}
\usepackage{caption}
\DeclareCaptionFormat{custom}{{\small\textbf{#1#2}\textit{#3}}}
\captionsetup{format=custom}

\usepackage{showyourwork}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{wasysym}
\usepackage{xcolor}
\usepackage{nicefrac}
\usepackage{multirow}
\usepackage[affil-it]{authblk}
\usepackage{graphicx}
\usepackage{biblatex}
\addbibresource{bib.bib}

@@

            We extend a previously published \gls{hmm}, which models the \glspl{lnl} (I, II, III, IV, V, and VII) as hidden binary \glspl{rv}. Each represents a patient's true state of lymphatic involvement. Clinical involvement at diagnosis represents the observed binary \glspl{rv} linked to the true state via sensitivity and specificity. The primary tumor and the hidden \glspl{rv} are connected in a graph. Each edge represents the conditional probability of metastatic spread per abstract time-step, given disease at the edge's starting node. To learn these probabilities, we draw \acrlong{mcmc} samples from the likelihood of a dataset (\variable{output/num_patients.txt} \gls{opscc} patients) from three institutions. We compute the model evidence using \acrlong{ti} for different graphs to determine which describes the data best.

@@

\begin{table}[!h]
\centering
\begin{tabular}{|cc|cc|}
    \hline
    & & $X_2=0$ & $X_2=1$ \\
    \hline
    \multirow{2}{*}{$X_3$} & $=0$ & $1-b_3$ & $(1-b_3)(1-t_{23})$ \\
    & $=1$ & $b_3$ & $1-b_3-t_{23}+b_3 t_{23}$ \\
    \hline
\end{tabular}
% \caption{$P\left( X_3 \mid X_2 \right)$ for all four possible combinations of the states of $X_3$ and $X_2$.}
\end{table}

@@

\begin{table}[!h]
\centering
\begin{tabular}{|cc|cc|}
    \hline
    & & $X=0$ & $X=1$ \\
    \hline
    \multirow{2}{*}{$Z$} & $=0$ & $s_P$ & $1-s_N$ \\
    & $=1$ & $1-s_P$ & $s_N$ \\
    \hline
\end{tabular}
% \caption{$P\left( Z \mid X \right)$ for all possible values of $X$ and $Z$.}
\end{table}

@@

\begin{figure}[!b]
    \centering
    \includegraphics[width=0.6\linewidth]{figures/extended_graph.png}
    \caption{Extended \gls{dag} representing different possible spread graphs underlying the \gls{hmm}. As in \cref{fig:graph_with_obs}, red arcs are parametrized with probabilities of spread from the tumor (red circle) to the \glspl{lnl} (blue circles). These red arcs, together with the blue arcs fron \gls{lnl} to \gls{lnl}, make up the \emph{base graph}. One after the other, each of the green arcs was added to the base graph. Subsequently, the performance of the resulting models in terms of its \gls{bic} was compared to the base graph to assess whether the additional edge should be kept in the \emph{winning graph} or not. \label{fig:extended_graph}}
\end{figure}

@@

\begin{figure}[t]
    \script{thermo_int.py}
    \centering
    \includegraphics[width=\linewidth]{figures/thermo_int.png}
    \caption{The expected log-accuracy under the power posterior plotted against the value of the inverse temperature $\beta$. These values were computed during a TI run for the loosing graph model (red line), as well as the winning graph model (green line). Note that the scaling of the x-axis is chosen such that the 64 points on the temperature ladder appear evenly spaced. This is to stress how the log-accuracy develops in the range from $\beta=0$ to around $\beta=0.1$. The area under these two curves yields an approximation of the log-evidence of the respective model. It is barely visible that due to the simplicity of the loosing graph model its accuracy increases earlier and faster that the winning graph model's. However, already before $\beta=0.031$ the model based on the winning graph outperforms the loosing graph model. Ultimately, the winning graph's log-evidence is better than the loosing graph model's.}
    \label{fig:thermo_int}
\end{figure}

@@

\begin{table}[b]
\centering
\begin{tabular}{|c|c|c|}
    \hline
    $K_\text{1v2}$ & $\ln{K_\text{1v2}}$ & support for $\mathcal{M}_1$ \\
    \hline
    $< 10^0$ & $< 0$ & negative (supports $\mathcal{M}_2$) \\
    $10^0$ to $10^{\nicefrac{1}{2}}$ & 0 to 1.15 & barely worth a mention \\
    $10^{\nicefrac{1}{2}}$ to $10^1$ & 1.15 to 2.3 & substantial \\
    $10^1$ to $10^{\nicefrac{3}{2}}$ & 2.3 to 3.45 & strong \\
    $10^{\nicefrac{3}{2}}$ to $10^2$ & 3.45 to 4.6 & very strong \\
    $> 10^2$ & $> 4.6$ & decisive \\
    \hline
\end{tabular}
\caption{Interpretation of Bayes factors and their natural logarithms in terms of their support for or against one of the two compared models as introduced by \cite{jeffreys_theory_1998}.}
\label{table:bayes_factor}
\end{table}

@@

\begin{table}[t]
\centering
\begin{tabular}{|l|rr|}
    \hline
    \textbf{Modality} & \textbf{Specificity} & \textbf{Sensitivity} \\
    \hline
    \acrshort{ct} & 76\% & 81\% \\
    \acrshort{pet} & 86\% & 79\% \\
    \gls{mri} & 63\% & 81\% \\
    \acrshort{fna} & 98\% & 80\% \\
    Pathology & $\approx$ 100\% & $\approx$ 100\% \\
    \hline
\end{tabular}
\caption{Literature sensitivity and specificity values that we used to infer the most likely involvement for a patient when multiple diagnostic modalities reported conflicting nodal involvement \cite{de_bondt_detection_2007,kyzas_18f-fluorodeoxyglucose_2008}.}
\label{table:sens_spec}
\end{table}

@@

\begin{table}[t]
\centering
\begin{tabular}{|ccccccrrrr|}
    \hline
    \multicolumn{6}{|c}{\textbf{LNL involvement} } & \multicolumn{4}{c|}{\textbf{T-category} } \\
    I & II & III & IV & V & VII & \multicolumn{2}{c}{\textbf{early} } & \multicolumn{2}{c|}{\textbf{advanced} } \\
    \hline
    \variable{output/data_table.tex} \\
    \hline
\end{tabular}
\caption{Prevalence of involvement patterns in the multi-centric dataset. An involvement pattern is characterized by the state of the six \glspl{lnl}: A red dot means the \gls{lnl} was reported to be metastatic, a green dot means it was determined to be healthy and a question mark means that the prevalence was marginalized over the state of this \gls{lnl}.}
\label{table:data_prevalence}
\end{table}

@@

\begin{figure*}[t]
    \centering
    \includegraphics[width=\linewidth]{figures/ranking.pdf}
    \caption{Visual ranking of the investigated graphs w.r.t. their model evidence, computed via thermodynamic integration. Not shown are the two graphs where the arc from \gls{lnl} III to IV was removed. Their respective model evidence fell below $-1528$ and the two graphs would appear far left in the figure. In the bottom left corner, we provide a visual reference in analogy to \cref{table:bayes_factor}: E.g., any difference in the model evidence shorter than the first of the three rulers indicates that the improvement is ``barely worth a mention''. \label{fig:ranking}}
\end{figure*}

@@

@@

\begin{figure}[!b]
    \script{wg_III_risks.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/wg_III_risks.png}
        \caption{Histograms over the risk for microscopic involvement in \gls{lnl} III, given that a patient presents as clinically N0 (green), or given that the patient's \gls{lnl} II shows clinical involvement (blue). The top panel displays these risks for early T-category, the bottom panel for advanced T-category.}
        \label{fig:wg_III_risks}
    \end{centering}
\end{figure}

@@

@@

\subsection{Implications for Elective\\Nodal Treatment}

Risk predictions obtained by the model may be used to design clinical trials on volume-deescalated treatment of \gls{opscc}. In the context of radiotherapy, this corresponds to excluding \glspl{lnl} from the \gls{ctv-n}, which are irradiated according to the current guidelines. The list below should be seen as a summary of \cref{subsec:results:risk_prediction} and the limitations discussed in \cref{subsec:disc:limitation} should be taken into account in its interpretation. Assuming that one accepts approximately a 5\% risk of occult metastases per \gls{lnl}, the statistical model presented in this paper would suggest to:
\begin{itemize}
    \item[$\bullet$] Irradiate level II for all patients.
    \item[$\bullet$] Irradiate level III for most patients. Only for clinically N0, early T-category patients, not irradiating level III could be considered. 
    \item[$\bullet$] Exclude level IV from the \gls{ctv-n} for patients with clinically negative level III. For advanced T-category patients with involvement of level III, level IV should be irradiated.
    \item[$\bullet$] Exclude level V from the \gls{ctv-n} for most patients. Only for patients with extensive involvement of levels II, III, and IV, irradiation of level V can be considered.
    \item[$\bullet$] Exclude level I from the \gls{ctv-n} for early T-category patients with limited metastatic disease. For advanced T-category patients with extensive nodal involvement of levels II, III, and IV, level I should be irradiated (see also the limitations discussed in \cref{subsec:disc:limitation}).
    \item[$\bullet$] Exclude level VII in all patients, unless it is clinically involved.
\end{itemize}

@@

\begin{figure}[!hb]
    \script{wg_I_risks.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/wg_I_risks.png}
        \caption{Distributions over predicted risk for involvement in \gls{lnl} I, given different clinical diagnosis scenarios: For N0 patients (green), patients with macroscopic involvement in \gls{lnl} II (blue), and for the case where the \glspl{lnl} II, III, and IV show involvement. The top panel shows these risks for early T-category and the bottom row for advanced T-category. \label{fig:wg_I_risks}}
    \end{centering}
\end{figure}

@@

\begin{figure}[!b]
    \script{wg_sens_spec_risks.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/wg_sens_spec_risks.png}
        \caption{Dependence of risk of occult disease on sensitivity and specificity. The figure shows the risk of involvement in level IV, given the mean of the parameter samples, drawn during sampling, in patients with clinical involvement of levels II and III. \label{fig:wg_sens_spec_risks}}
    \end{centering}
\end{figure}

The patient data detailing lymphatic involvement is publicly available in the form of CSV tables in the GitHub repository \href{https://github.com/rmnldwg/lydata}{\texttt{rmnldwg/lydata}}. In another GitHub repository, \href{https://github.com/rmnldwg/lynference}{\texttt{rmnldwg/lynference}}, we define the experiments based on different graph structures. These experiments are reproducible via the tool \href{https://dvc.org}{DVC} and their locally computed results are uploaded to an Azure blob storage container. Lastly, the \LaTeX\ source code, all Python scripts to generate the figures and tables in this work, and the pipeline definition to build the document (using \showyourwork \cite{luger_mapping_2021}) are made public in the GitHub repository \href{https://github.com/rmnldwg/graph-extension-paper}{\texttt{rmnldwg/graph-extension-paper}}. To rebuild it, the \showyourwork tool pulls the patient data from the \href{https://github.com/rmnldwg/lynference}{\texttt{rmnldwg/lynference}} repository and the trained models in the form of \gls{mcmc} samples from the Azure blob storage container, then runs the Python scripts and compiles the \LaTeX\ article. Detailed instructions on how to reproduce the entire pipeline, including the individual experiments, are gladly given upon request. They may also be found in the respective repository's \texttt{README.md} file.
